---
title: "qtl2fst user guide"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Create fst_genoprob for qtl2 data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 7, fig.height = 5)
```

Memory usage can be a big obstacle in the use of
[R/qtl2](http://kbroman.org/qtl2), particularly regarding the QTL
genotype probabilities calculated by `calc_genoprob()`. For dense
markers in multi-parent populations, these can use gigabytes of RAM.

This led us to develop ways to store the genotype probabilities on
disk. In the present package, we rely on the
[fst package](http://www.fstpackage.org), which includes the option to
compress the data.

Let's first load the R/qtl2 and R/qtl2fst packages.

```{r, load_packages}
library(qtl2)
library(qtl2fst)
```

In this vignette, we'll give a quick illustration of the
[R/qtl2fst](https://github.com/rqtl/qtl2fst) package using the
[iron
dataset](https://kbroman.org/qtl2/pages/sampledata.html#f2-intercross)
included with [R/qtl2](https://kbroman.org/qtl2).
We'll first load the data.

```{r, load_iron_data}
iron <- read_cross2(system.file("extdata", "iron.zip", package="qtl2"))
```

Let's calculate the genotype probabilities and convert them to allele probabilities.

```{r, calc_alleleprob}
pr <- calc_genoprob(iron, error_prob=0.002)
apr <- genoprob_to_alleleprob(pr)
```

Use the function `fst_genoprob()` to write the probabilities to a fst database.
You could do the same thing with the allele probabilities.

```{r write_fst_db}
tmpdir <- file.path(tempdir(), "iron_genoprob")
dir.create(tmpdir)
fpr <- fst_genoprob(pr, "pr", tmpdir)
fapr <- fst_genoprob(apr, "apr", tmpdir)
```

The genotype probabilities are saved in a set of files, one per
chromosome. There is also an RDS index file, which is a copy of the
index object returned by `fst_genoprob()`.

```{r list_files}
list.files(tmpdir)
```

You can treat the `fpr` and `fapr` objects as if they were the
genotype probabilities themselves. For example, use `names()` to get
the chromosome names.

```{r, names_fpr}
names(fpr)
```

### Selecting one chromosome

If you selecting a chromosome, it will be read from the fst database
and into an array.

```{r, read_chromosome}
apr_X <- fapr[["X"]]
dim(apr_X)
```

You can also use the `$` operator.

```{r}
apr_X <- fapr$X
dim(apr_X)
```

### Subsetting by ind, chr, mar

You can subset by individuals, chromosome, and markers, with
`subset(object,ind,chr,mar)` or `[ind,chr,mar]`. Just the selected
portion will be read, and the fst database will not be altered.

```{r subset_fapr}
selected_ind <- subset(fapr, ind=1:20, chr=c("2","3"))
dim(fapr)
```

You can also subset with brackets in various ways.

```{r subset_brackets}
fapr_sub1 <- fapr[1:20, c("2","3")][["3"]]
fapr_sub2 <- fapr[,"2"]
fapr_sub23 <- fapr[,c("2","3")]
fapr_subX <- fapr[,"X"]
```

You can use a third dimension for markers, but be careful that if you
select a subset of markers that excludes one or more chromosomes,
those will be dropped.

```{r select_markers}
dim(subset(fapr, mar=1:30))
dim(fapr[ , , dimnames(fapr)$mar$X[1:2]])
```

### Binding by columns or rows.

Binding by columns (chromosomes) or rows (individuals) may cause
creation of a new fst database if input objects arose from different
fst databases. However, if objects are subsets of the same
`"fst_genoprob"` object, then it reuses the one fst database. Further,
if objects have the same directory and file basename for their fst
databases, they will be combined without creation of any new fst
databases.

See `example(cbind.fst_genoprob)` and `example(rbind.fst_genoprob)`
with objects having distinct fst databases.

Here's column bind (chromosomes).

```{r cbind_fapr}
fapr_sub223 <- cbind(fapr_sub2,fapr_sub23)
```

And here's row bind (individuals)..

```{r rbind_fapr}
f23a <- fapr[1:20, c("2","3")]
f23b <- fapr[40:79, c("2","3")]
f23 <- rbind(f23a, f23b)
```

Subset on markers. This way only extracts the selected `markers` from
the fst database before creating the array.

```{r subset_markers}
markers <- dimnames(fapr$X)[[3]][1:2]
dim(fapr[,,markers]$X)
```

This way extracts all markers on `X`, creates the array, then subsets on selected `markers`.

```{r extract_markers_chr_X}
markers <- dimnames(fapr$X)[[3]]
dim(fapr$X[,,markers[1:2]])
```

Two `"fst_genoprob"` objects using the same database. Combine using `cbind`. Notice that the order of chromosomes is reversed by joining `fapr2` to `fapr3`. Be sure to not overwrite existing fst databases!

```{r cbind_two_subsets}
fapr2 <- fst_genoprob(subset(apr, chr="2"), "aprx", tmpdir)
fapr3 <- fst_genoprob(subset(apr, chr="3"), "aprx", tmpdir)
fapr32 <- cbind(fapr3,fapr2)
dim(fapr32)
list.files(tmpdir)
```

### Looking under the hood at `"fst_genoprob"` object

Here are the names of elements in a `"fst_genoprob"` object:

```{r names_fapr}
names(unclass(fapr))
```

```{r fapr_fst_path}
unclass(fapr)$fst
```

```{r ind_chr_mar_pieces}
sapply(unclass(fapr)[c("ind","chr","mar")], length)
```

An `"fst_genoprob"` object has all the original information. Thus, it
is possible to restore the original object from a `subset` (but not
necessarily from a `cbind` or `rbind`). Here is an example.

```{r restore_from_subset}
fapr23 <- subset(fapr, chr=c("2","3"))
dim(fapr23)
dim(fst_genoprob_restore(fapr23))
```

### Paths

Use `fst_path()` to determine the path to the fst database.

```{r path_to_fpr}
fst_path(fpr)
```

If you move the fst database, or if it's using a relative path and you
want to work with it from a different directory, use `replace_path()`.

```{r new_path_to_fpr}
fpr_newpath <- replace_path(fpr, tempdir())
```

### Direct construction of the fst database

Since the genotype probabilities can be really large, it's very RAM
intensive to calculate all of them and then create the database.
Instead, you can use `calc_genoprob_fst()` to run `calc_genoprob()`
and then `fst_genoprob()` for one chromosome at a time.

```{r calc_genoprob_fst}
fpr <- calc_genoprob_fst(iron, "pr", tmpdir, error_prob=0.002, overwrite=TRUE)
```

Similarly, `genoprob_to_alleleprob_fst()` will run
`genoprob_to_alleleprob()` and then `fst_genoprob()` for one
chromosome at a time.

```{r genoprob_to_alleleprob_fst}
fapr <- genoprob_to_alleleprob_fst(pr, "apr", tmpdir, overwrite=TRUE)
```

### Genome scans

You can use the `fst_genoprob()` object in place of the genotype
probabilities, in genome scans with `scan1()`.

```{r genome_scan}
Xcovar <- get_x_covar(iron)
scan_pr <- scan1(fpr, iron$pheno, Xcovar=Xcovar)
find_peaks(scan_pr, iron$pmap, threshold=4)
```


Similarly for calculating QTL coefficients with `scan1coef()` or
scan1blup()`:

```{r coef}
coef16 <- scan1coef(fpr[,"16"], iron$pheno[,1])
blup16 <- scan1blup(fpr[,"16"], iron$pheno[,1])
```


```{r clean_up_files, include=FALSE}
unlink(tmpdir)
```
